---
title: 'Statistical Inference: Course Project - Part 1'
author: "Nikita Kirnosov"
output:
  pdf_document:
    toc: yes
  html_document:
    css: custom.css
    keep_md: yes
    toc: yes
---
```{r custon_knitr, echo=FALSE}
library(knitr)
library(xtable)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knit_hooks$set(inline = function(x) {
   if (is.numeric(x)) round(x, 2)})
options(digits=2)
```

## Overview

In this project we will investigate the exponential distribution in R and compare it 
with the Central Limit Theorem. The exponential distribution can be simulated in R 
with $rexp(n, \lambda)$ where $\lambda$ is the rate parameter. The mean of exponential 
distribution is $1/\lambda$ and the standard deviation is also $1/\lambda$. 

We will perform a thousand simulations to investigate the distribution of averages 
of 40 exponentials with $\lambda$ set to 0.2. The validity of the CLT will be evaluated
and discussed.

## Simulations
 
 The seed was set to 111 to make the results reproducible.

```{r}
set.seed(111)
lambda <- 0.2
num_sim <- 1000
sample_size <- 40
```

Since we only using means of each simulated normal distribution,
we will simulate `r num_sim*sample_size` numbers and store them 
in a matrix with `r num_sim` columns and `r sample_size` rows.
Taking the mean of the rows will leave us with a vector of sample 
means.

```{r}
means <- rowMeans(matrix(rexp(num_sim*sample_size, rate=lambda), num_sim, sample_size))
```

\newpage 

## Results

The histogram below overlays the distribution of samples' means and its mean
(black dashed line) with the theoretically predicted shape and mean (in red).

```{r echo=FALSE}
hist(means, breaks=sample_size, prob=TRUE,
     main=expression(paste("Samples' averages distribution (",lambda,"=0.2)")),
     xlab="")
abline(v=mean(means),lty=2, lwd=2, col="black")
abline(v=1/lambda,lty=2, lwd=2, col="red")
xfit <- seq(min(means), max(means), length=num_sim)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(sample_size)))
lines(xfit, yfit, pch=22, col="red")
legend('topright', c("simulation", "theoretical"), lty=c(1,1), col=c("black", "red"))
```

Aside from the visual examination, the comparison of these distrubutions' 
mean and variance values reveals that they ar very close, thus supporting
the hypothesis of distribution normality:

```{r kable, echo=FALSE}
 x <- rbind(c("","Simulated","Theoretical"),
            c("mean",signif(mean(means),3),"5.00"),
            c("variance",signif(var(means),3),signif(1/(lambda^2 * sample_size),3))) 
kable(x)
```

Another aid in discussion of sample normality is the normal Q-Q plot.
The linearity of the points suggests that the data are normally distributed.

```{r, echo=FALSE}
qqnorm(means,col="blue")
qqline(means)
```

## Conclusion

The obtained data suggests that the Central Limit Theorem is applicable 
in this case. This assumption can be also confirmed by Kolmogorov-Smirnov
normality test, where high p-value would suggest the normal distribution:

```{r}
ks.test(means,"pnorm",mean(means),sqrt(var(means)))
```

\newpage

## Appendix

Plain R code used in this project:

```{r eval=FALSE}
library(knitr)
library(xtable)
opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knit_hooks$set(inline = function(x) {
   if (is.numeric(x)) round(x, 2)})
options(digits=2)

set.seed(111)
lambda <- 0.2
num_sim <- 1000
sample_size <- 40

means <- rowMeans(matrix(rexp(num_sim*sample_size, rate=lambda), num_sim, sample_size))

hist(means, breaks=sample_size, prob=TRUE,
     main=expression(paste("Samples' averages distribution (",lambda,"=0.2)")),
     xlab="")
abline(v=mean(means),lty=2, lwd=2, col="black")
abline(v=1/lambda,lty=2, lwd=2, col="red")
xfit <- seq(min(means), max(means), length=num_sim)
yfit <- dnorm(xfit, mean=1/lambda, sd=(1/lambda/sqrt(sample_size)))
lines(xfit, yfit, pch=22, col="red")
legend('topright', c("simulation", "theoretical"), lty=c(1,1), col=c("black", "red"))

rbind(c("","Simulated","Theoretical"),
        c("mean",signif(mean(means),3),"5.00"),
        c("variance",signif(var(means),3),signif(1/(lambda^2 * sample_size),3))) 

qqnorm(means,col="blue")
qqline(means)

ks.test(means,"pnorm",mean(means),sqrt(var(means)))
```
