---
title: 'Statistical Inference: Course Project - Part 2'
author: "Nikita Kirnosov"
output:
  pdf_document:
    toc: yes
  html_document:
    css: custom.css
    keep_md: yes
    toc: yes
---
```{r custon_knitr, echo=FALSE}
library(knitr)
library(pander)
library(ggplot2)
library(datasets)

opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
knit_hooks$set(inline = function(x) {
   if (is.numeric(x)) round(x, 2)})
options(digits=2)
```

## Overview

The response is the length of odontoblasts (teeth) in each of 
10 guinea pigs at each of three dose levels of Vitamin C 
(0.5, 1, and 2 mg) with each of two delivery methods 
(orange juice or ascorbic acid) is under study in present work.
The resulting data set has the following characteristics:

```{r}
summary (ToothGrowth)
```

In current paper we aim to elucidate the dependences of the teeth 
length on 

* vitamin C dose,

* delivery method.

## Results

The graph below indicates for both methods of delivery the dose increase results
in teeth length inctrease. It also appears that for doses of 0.5 and 1.0 mg 
orange juice reponse is stronger than for ascorbic acid. Let us test this observation.

```{r}
ggplot(data=ToothGrowth, aes(x=as.factor(dose), y=len, fill=supp)) +
  geom_boxplot(outlier.colour = "black",outlier.size = 3,outlier.shape=3) +
  geom_dotplot(binaxis="y",binwidth = 1,stackdir="center",dotsize=0.25)+
  facet_grid(. ~ supp) +
  xlab("Dose, mg") +
  ylab("Tooth length, mm") +
  guides(fill=guide_legend(title="Delivery Method"))
```

To decide whether there is a real difference between the groups by
dose level and delivery method, let us a series of two-sided
unpaired t-tests to obtain the confidence intervals and p-values.
Here we will assume unequal variance, which is suggested by the graph. 
 
Our null hypothesis will be that there is no difference in the means
between the two groups. Let us see whether this is true.

```{r}
t_test <- function(x) {
    dt <- t.test(len ~ supp, data=subset(ToothGrowth, dose==x), 
                 paired=FALSE, var.equal=FALSE)
    return(c(sprintf("%.3f", round(x,1)),
             sprintf("%.3f", round(dt$stat,3)),
             sprintf("%.3f", round(dt$p.value,3)),
             paste("[",round(dt$conf.int[1],3),",",round(dt$conf.int[2],3),"]")))
    }

df <- data.frame (Name=c("dose, mg", "t", "p-value",
                         "95% confidence interval"),
                  Dose_1=t_test(0.5),Dose_2=t_test(1),Dose_3=t_test(2))
pander(df, split.tables=120,
       caption="*Comparison of delivery methods by dose*")
```

Table 1 presents t- and p-values for each vitamin C dose level as well as
corresponding 95% confidence intervals. In order to be able to decide whether 
different dose levels yield the results which are statistically significantly
different for different delivery methods, we should adjust the p-values.
Here we chose to use Holm's method:

```{r}
p_values <- c(as.numeric(as.character(df[3,2])),
       as.numeric(as.character(df[3,3])),
       as.numeric(as.character(df[3,4])))
```

```{r, echo=TRUE}
round(p.adjust(p_values, method = "holm"),3)
```

## Conclusion

* At the 0.5 and 1 mg dose levels, there is a statistically significant difference
  between the means of the two delivery options, since the adjusted p-values are significant
  at the $\alpha$ = 0.05 level, and the 95% confidence intervals do not include zero.

* The 2 mg dose level does not indicate the statistically significant difference in 
  delivery methods, but it has to be noted that could be a consequence of the 
  extremely low test power due to a small number of data points.
  
* The possible conclusion can be made that orange juice is a more efficient source
  of vitamin C for guinea pigs teeth growth. The probable explanation is that it 
  contains other nutrients used in this physiological process. It is also clear that
  the increase of the vitamin C dose from 0.5 mg to 1.0 mg results in positive
  response regardless of the supplementation method.

## Appendix

R code used:

```{r, eval=FALSE, echo=TRUE}
library(knitr)
library(pander)
library(ggplot2)
library(datasets)

opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
knit_hooks$set(inline = function(x) {
   if (is.numeric(x)) round(x, 2)})
options(digits=2)

summary (ToothGrowth)

ggplot(data=ToothGrowth, aes(x=as.factor(dose), y=len, fill=supp)) +
  geom_boxplot(outlier.colour = "black",outlier.size = 3,outlier.shape=3) +
  geom_dotplot(binaxis="y",binwidth = 1,stackdir="center",dotsize=0.25)+
  facet_grid(. ~ supp) +
  xlab("Dose, mg") +
  ylab("Tooth length, mm") +
  guides(fill=guide_legend(title="Delivery Method"))

t_test <- function(x) {
    dt <- t.test(len ~ supp, data=subset(ToothGrowth, dose==x), 
                 paired=FALSE, var.equal=FALSE)
    return(c(sprintf("%.3f", round(x,1)),
             sprintf("%.3f", round(dt$stat,3)),
             sprintf("%.3f", round(dt$p.value,3)),
             paste("[",round(dt$conf.int[1],3),",",round(dt$conf.int[2],3),"]")))
    }

df <- data.frame (Name=c("dose, mg", "t", "p-value",
                         "95% confidence interval"),
                  Dose_1=t_test(0.5),Dose_2=t_test(1),Dose_3=t_test(2))
pander(df, split.tables=120,
       caption="*Comparison of delivery methods by dose*")

x <- c(as.numeric(as.character(df[3,2])),
       as.numeric(as.character(df[3,3])),
       as.numeric(as.character(df[3,4])))

round(p.adjust(x, method = "bonferroni"),3)
```